Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2017-06-21 LatticeOptics/latticeOpticsV4.tex
new revision

\author{A.M., S.B., F.H., M.P., C.C., G.J., D.G., M.G.}
\author{M. Greiner}
\affiliation{
  Department of Physics, Harvard University,
  Cambridge, Massachusetts, 02138, USA}

\date{\today}
\begin{abstract}
We describe the design of a stable high power $1064\,$nm laser system for use in optical lattice experiments. The system is based on a low noise seed laser amplified with an array of four heavily modified \textit{Nufern} fiber amplifiers. The intensity of every beam is stabilized with a low noise, nonlinear feedback loop. The positional stability of the system is characterized and monitored in real time.  
\end{abstract}
\maketitle

Ultracold atoms trapped in an optical lattice has become a powerful platform for experiments ranging from investigations of strongly correlated systems, quantum state engineering, quantum computing architectures and precision metrology\cite{Ye}.
Further, the recent advent of quantum gas microscopy has enabled studies that exhibit unprecedented control over few atom systems at extremely low energy scales.
However, this rapid progress places ever more stringent technical constraints on the lasers used to trap and manipulate the atomic systems.
For example, heating due to the intensity noise of any laser cannot be allowed to approach the longest thermalization timescales, lest the system fail to thermalize~\cite{Savard1997}. 
Since thermalization timescales generally grow with falling temperature, this complicates attaining low temperatures.
Similarly, experiments that rely on a small number of atoms held by multiple traps places stringent requirements on the relative positional stability of the traps, to ensure that the operations performed by the traps have adequate fidelity.

To address these concerns, we discuss a high power optical lattice system that exhibits levels of intensity noise below $-120\,$dBc between $1\,$kHz and $3\,$MHz.
The positional stability of the lattice phase is at the $<0.5\mu$m level and the stability of the harmonic confinement is at the few site level, measured over the course of hours.

\begin{figure}
  \begin{center}
    \includegraphics{fig/figure1/figure1.pdf}
    \caption{\textbf{High level overview} showing the major components of the lattice laser feedback system.}
  \end{center}
\end{figure}

The architecture is based on a single low-power, low-noise seed laser that supplies light to an array of four high-power fiber amplifiers (one for a transport beam~\cite{Huber2014}, and one each for the $X,Y,Z$ lattices).
These amplifiers generate enough light to perform site resolved imaging~\cite{Parsons2016}, but also substantially increase the level of noise and degrade the mode shape. 
The fiber amplifier output is controlled by two feedback loops, which stabilize the light in the low power and high power regimes.
After stabilization, the laser is mode-shaped to produce a desired waist when focused in the atom plane. 
Just before the beam is launched toward the atoms, a small sample of it is imaged by a camera to monitor the pointing.

\section{Laser system}
The lattice laser system is seeded by an Innolight Mephisto laser, which provides approximately 2 W of 1064 nm light. 
This commercial laser utilizes a non-planar ring oscillator (NPRO) design that ensures superb phase and intensity stability. This source is split into four fibers (Thorlabs PM980-XP) using polarizing beam splitters (PBS).
To avoid interference of the lattice beams, two of the beam paths pass through acousto-optic modulators (AOM), which detunes the lattice frequencies from each other. Figure \ref{fig:noises} shows the noise performance of only this subsystem.

Quantum gas microscopy of \Li{} requires single site trap frequencies of approximately 1 MHz~\cite{Parsons2015} in order to be in the Lamb-Dicke regime, which is necessary to perform Raman sideband cooling.
Such trap frequencies may only be attained using tightly focused high power beams, which requires that the seed power be amplified to approximately 30 W per lattice axis, which is done using Nufern Fiber amplifiers (NuAmp SUB1174-34).
These amplifiers can amplify 150 mW of seed light to up to approximately 45 W. They use a length of doped fiber pumped with 808 nm light from fiber coupled diode sources to amplify 1064 nm light. 
The amplification is performed using a two stage amplifier design, where the stages are controlled using separate power supplies and electronics.
The high optical powers supplied by the fiber amplifiers raise two engineering difficulties:
\begin{enumerate}
  \item The fiber amplification process is susceptible to many technical noise sources, such as acoustic coupling of the fiber to the environment and electrical design of the amplifier, 
  \item Stimulates Brillouin scattering (SBS) places a hard lower bound on the noise floor~\cite{Agrawal2013}. 
\end{enumerate}

The NuAmp suffers from several problems: it is controlled via a USB interface, which is suspected to cause ground loops which add noise and interference to the analog systems, as seen in figure \ref{fig:noises}.
We solve this problem by replacing the USB control board with a custom-designed, Ethernet based solution that is compatible with our custom control system and that carefully avoids ground loops. 
There are two power supplies in the fiber amplifier - a general purpose supply for the control electronics and first amplification stage and a high power current supply for the second stage. 
Despite the presence of a water-cooled cold-plate to cool the optical fiber and pump diodes, the high power supply is air cooled using a fan approximately $20$ cm away from the gain fiber. 
This fan is suspected to lead to increased noise at acoustic frequencies.
In order to reduce acoustic coupling and reduce switching power supply noise, both power supplies were removed from the fiber amplifier chassis, and the low power supply was replaced by a linear PSU (Acopian A24MT350M). 
Further, a high current line filter (MPE DS26387) was added to the high-power PSU (Lumina LDD-600-60-10-5VP/M).
Combined, these changes serve to suppress many of the noise spurs measured in the RIN of the fiber amplifier, as seen in figure \ref{fig:noises}. 
The measurements in parts (a)-(d) were carried out on a single fiber amplifier, prior to and after the modifications. The initial and final noise spectra were verified to be consistent between amplifiers coming from the same manufacturing batch.
The figure is based on fiber amplifiers from a relatively new batch, which, even in its unmodified configuration does not exhibit a large switching supply spur. 
This is not the case for certain older batches, where the spur is approximately 30 dB higher than the background.

The operation of the Nufern fiber amplifiers can be further optimized by finely controlling the cooling water required, which by tunes the wavelength of the light supplied by the diodes that optically pump the gain fiber.
Optimizing the temperature is advantageous because (1) the maximum output power of the fiber amplifier is increased due to the improved efficiency of the pump light and (2) the lifetime of the device is increased because less stray pump light needs to be absorbed when the pump light is separated from the desired light, reducing the thermal load on the system.

This is possible because the pump diodes are fixed directly to the water-cooling plate without additional temperature regulation.
Adding a Peltier TEC and regulating to the optimal wavelength is possible, but is technically difficult as it requires large cooling power of $\approx100$ W per each of the three pump diodes and carries the associated complexity of three feedback loops.
We use a single, dedicated heat exchanger (TermoTek P21518970) to supply cooling water to an array of four fiber amplifiers and tune the temperature such that the average power supplied by the amplifiers supplying the $X$ and $Y$ lattices is optimized, while simultaneously keeping the temperature of the amplifier well within a safe range.
Experimentally, we have found that more power can be gained by running the amplifiers at hotter temperature, but we have avoided doing this in an effort to increase the lifetime.
We have experimentally found that there is variation in the sensitivity to the cooling water temperature between fiber amplifiers. 
%Figure\ref{fig:pow_and_noise}c. shows the responses of the fiber amplifiers currently in the system.

@@

@@

The free-space propagation of the fiber amplifier output presents several technical challenges:
\begin{enumerate}
  \item The high optical power poses a significant danger to users and equipment.
  \item The high optical power is sufficient to induce substantial thermal lensing in some optics. 
  \item The positional fluctuations of the waist at the atoms cannot be allowed to exceed a few microns from shot to shot. 
\item By construction, the beam is retro-reflected, threatening the lifetime of the amplifier unless sufficient isolation is used. 
\item Approximately $1$ W of power leaks into the undesired cladding modes of the fiber. 
  \end{enumerate}

In order to address these challenges, we carefully engineered an optical system, shown in figure \ref{fig:optical_layout}, that cleans up the beam, controls its intensity between $10\,\mu$W and $20$ W, and ensures good pointing stability.
Due to the high laser power, fused silica glass is used wherever possible.
%except in the acousto-optic modulator (which is TeO$_2$), Berek compensator (which is quartz), and optical isolators, to minimze thermal lensing.
For the same reason, IBS coatings are used wherever possible, due to the higher damage thresholds.
Further, even small reflections pose both a danger to the users and risk damaging cables or starting a fire, so to alleviate this danger, all undesired beams above $100$ mW are directed to water cooled beam-dumps, which are placed as far as possible from critical beam paths. 
Weaker beams are dumped on uncooled diffusive beam catchers.

The large-diameter fiber tip at the output of the fiber amplifier is mounted in a monolithic mount made from oxygen-depleted copper.
The beam is collimated using an $f=20$ mm fused-silica, air-spaced triplet collimator (Opto-Sigma HFTLSQ-15-20PF1).
Then, the polarization is cleaned up using an IBS coated thin-film plate polarizer (Precision photonics PO1056-FY), which has the added benefit of rejecting the undesired cladding modes. 
The fact that the lattice is retro-reflected, combined with the observation that optical isolation falls with applied optical power~\cite{Yoshida1999} means that two stages of optical isolators must be used. 
The optical isolators are based on $5$ mm diameter isolators from Thorlabs (IO-5-1064-VHP), with the default cube polarizers replaced by IBS coated thin-film plate polarizers. 
Since the isolators are located approximately $1$ m away from the ultracold atomic system, undesired stray fields can have a dramatic effect.
To overcome this problem the isolators are enclosed in mu-metal shielding.

@@

After the beam has been cleaned up and isolated, it is important to consider two aspects of the desired experiment: 
\begin{enumerate}
\item The lattice power must be continuously tunable from around $10$ $\mu$W to around $20$ W. 
\item The location of the minimum Gaussian beam waist must be accurately positioned to overlap the other lattices and dipole traps present in the experiment, and must remain there for the duration of the collection process (hours to days).
\end{enumerate}
To satisfy the first of these requirements, a two-stage feedback system is implemented. The reason for the two stages is simple: the experiment typically operates in one of two regimes, which we will term \textit{detection} and \textit{interaction}. 
In the interaction phase of the experiment, the lattice is relatively shallow ($\approx 100-200$mW, corresponding to depths of $\approx 10 E_R$, where $E_R$ is the geometric recoil of the lattice), allowing atoms to tunnel and interact with other atoms. 
In this regime, fine and potentially fast control is required.
In the detection phase, the depth of the lattice is dramatically raised ($\approx 2000 E_R$), isolating atoms in their individual wells so that Raman sideband imaging can be performed~\cite{Parsons2015}.
In the detection regime, the control need not be fast, and the passive stability of the Mephisto laser ensures that noise is low at relevant frequency scales ($\approx 2$ MHz).
Thus, two loops are utilized, a fast loop that uses an acousto-optic modulator to actuate the laser power at low powers, and a slow loop which uses a Berek compensator, to actuate in the high power mode.

After the isolation stage, we use the sequence of optical elements shown in figure \ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on beam path) half-wave plate, and polarizer. 
The Berek compensator is a simple $z$-cut quartz plate coated to be anti-reflecting at 1064 nm. 
By tilting the plate about its vertical axis, the extraordinary axis is mixed into the propagation of the beam, leading to a tilt-dependent bi-refringence that can be tuned from a zero-wave plate past a half-wave plate.
Combined with the subsequent polarization optics, rotation of the Berek compensator changes the transmitted power with perfect contrast and transmitted power as a function of the tilt angle is shown in figure \ref{fig:berek}.
\begin{figure}
  \begin{center}
    \includegraphics{fig/berek_sim.pdf}
    \caption{\textbf{Berek compensator} performance as a function of the Berek tilt angle $\phi$ and waveplate tilt angles.}\label{fig:berek}
  \end{center}
\end{figure}
Full modulation contrast is achieved when the fast axes of the waveplates around the Berek compensator $\theta$ are at $22.5^\circ$ to the rotation axis of the Berek compensator.
In that case, if the Berek compensator does not change the polarization, the light is fully transmitted. 
If it functions as a half-wave plate, the light is fully rejected, with a smooth crossover between the two cases as the Berek angle, $\phi$, is adjusted.
This contrast can be reduced by varying $\theta$, to the limiting case where $\theta=0$, where the rotation of the Berek compensator does nothing, since the polarization of the light is parallel to the rotation axis of the quartz.

\begin{figure*}
  \begin{center}
    \includegraphics{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop} performance showing the impulse response (a) at relatively low powers and the corresponding Bode plots (b-c).}\label{fig:berek_step_response}
  \end{center}
\end{figure*}
A key point is that, as seen in figure \ref{fig:berek}, if the two waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator. 
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in figure \ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate. 
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.

\subsection{Low power feedback}
%The high power feedback used in the \textit{imaging} phase of the experiment and is exceedingly robust
%, and allows the user to raise the power and ensure an adequate Lamb-Dicke parameter\cite{Parsons2015}. 
The high power feedback used in the \textit{imaging} phase of the experiment is exceedingly robust because the heating rate from scattered resonant imaging light far exceeds any technical heating from the loop.
Even if the loop was exceedingly noisy, the heating would be entirely compensated by the simultaneous Raman cooling~\cite{Parsons2015}.

In contrast, the low power loop that controls the lattice laser during the \textit{interaction} phase of the experiment is critically important:
\begin{enumerate}
  \item Its RIN is one of the factors limiting the system temperature.
  \item It cannot affect the pointing.
  \item It must have closed loop bandwidths of $\geq1\mathrm{\,kHz}$, and open loop modulation bandwidths $\geq 100\mathrm{\,kHz}$.
\end{enumerate}
To this end, we use actuate the optical power using a TeO$_2$ AOM (Gooch \& Housego AOMO 3080-198), where the power of the RF supplied to the AOM is used to actuate the fast, low power feedback loop.
The AOM was chosen for its low thermal lensing and easily-accessible RF powers. 
It is mounted on a custom-built, monolithic, water-cooled flexure mount that allows optimization of the AOM efficiency, that has been found to remain stable over years of operation.

\begin{figure}
  \begin{center}
    \includegraphics{fig/circuits.pdf}
    \caption{\textbf{Feedback circuits.} (a) The reverse-biased photodiode. (b) Externally referenced phase-locked loop. (c) Nonlinear loop filter. (d) RF power modulation using a mixer.}\label{fig:circuits}
  \end{center}
\end{figure}
The feedback circuit is shown in figure \ref{fig:circuits}a. A pair of low noise photodiodes measures the optical power of the beam sampled with a weakly reflective optic (Thorlabs BSF10-C). 
The beam sampler is wedged in such a way that the two reflections diverge from one another, allowing us to illuminate both photodiodes. 
One of the sampled beams is attenuated by a factor of $20\times$ by reflection off of an uncoated fused silica glass plate, such that the dynamic range of its photodiode covers $0-20$ W of the beam directed at the atoms, and controls the Berek compensator based, high-power, feedback loop. 
The other beam is not attenuated, save by the beam sampler, and its dynamic range covers $0-1$ W.
To prevent damage to this photodiode when the system is at its highest powers, a shutter (Stanford Research Systems SR475) blocks the light when the high power loop is active.
The low power photodiode passes its signal $V_{\mathrm{pd}}$ to a loop filter circuit, which compares it to the control system's set-point, generated by digital to analog converter.
The loop filter's output controls the IF port of a microwave mixer (Minicircuits ZFM-2-S+), effectively modulating the RF power supplied by a low noise phase-locked loop (PLL).
The resulting RF signal is then amplified and sent to the AOM.
Let us now consider every part of this signal chain individually.

\subsubsection{Phase-locked loop (PLL)}
The local oscillator (LO) RF source is a custom-built printed circuit board (PCB) implementing a phase locked loop (PLL) based on the ADF4002 IC and Crystek's CVCO55CL-0060-0110 voltage controlled oscillator (VCO), preamplified with a low noise amplifier (Minicircuits PSA4-5043+). 
To ensure frequency stability (which translates directly into beam pointing of the laser), the PLL is locked to the $10$ MHz clock distributed around the lab.
In addition to PLL functionality, the board also contains two, digitally-controllable, high-isolation RF switches, giving a total isolation exceeding $90$ dB.

The spectrum and relative intensity noise of the PLL can be found in figure \ref{fig:pll}. 
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of typical VCO, without the slow frequency drift characteristic to these devices. 
Elimination of the carrier drift is worth the added phase noise because any drift of the frequency affects the steering of the laser beam passing through the AOM, which is entirely unacceptable since it would misalign the optical lattice.

>>>>

\author{A.M., S.B., F.H., M.F.P., C.S.C., G.J., D.G.}
\author{M. G.}
\affiliation{
  Department of Physics, Harvard University,
  Cambridge, Massachusetts, 02138, USA}

\date{\today}
\begin{abstract}
  We describe the design of a stable high power $1064\,$nm laser system for use in optical lattice experiments. The system is based on a low noise laser amplified by an array of four heavily modified, high-power fiber amplifiers. The intensity of every beam is stabilized using a nonlinear feedback loop. The position of the produced optical lattice is monitored in real time and its stability is characterized\footnote{This work is adapted from the PhD thesis of A.Mazurenko}.
\end{abstract}
\maketitle

Optical lattice holding an array of ultracold atoms have become a powerful experimental platform for a wide variety of investigations, including strongly correlated systems, quantum state engineering, quantum computing and precision metrology.
Further, the recent advent of quantum gas microscopy has enabled studies that exhibit unprecedented control over few atom systems at extremely low energy scales.
However, this progress has placed ever more stringent technical constraints on the lasers used to trap and manipulate the atomic systems.
For example, heating due to the intensity noise cannot be allowed to approach the longest thermalization timescales, lest the system fail to thermalize~\cite{Savard1997}.
Even if the system thermalizes, heating rates can limit the interrogation time available to probe a low-temperature state.
In a similar vein, experiments that rely on a small number of atoms held by multiple traps place stringent requirements on the relative positional stability of the traps, to ensure that the operations performed by the traps have adequate fidelity~\cite{Mazurenko2016}.

To address these concerns, we discuss a high power optical lattice system that exhibits levels of intensity noise below $-120\,$dBc between $1\,$kHz and $3\,$MHz.
The stability of the lattice phase is below the single site level and the stability of the harmonic confinement is at the few site level.

\begin{figure}
  \begin{center}
    \includegraphics{fig/figure1/figure1.pdf}
    \caption{\textbf{High level overview} showing the major components of the lattice laser feedback system.}\label{fig:high_level}
  \end{center}
\end{figure}

The coarse architecture of the system is shown in figure \ref{fig:high_level} and is based on a single low-power, low-noise seed laser that supplies light to an array of four high-power fiber amplifiers (one for a transport beam~\cite{Huber2014}, and one each for the $X,Y,Z$ lattices~\cite{Parsons2016}).
These amplifiers generate enough light to perform site resolved imaging~\cite{Parsons2015}, but also substantially increase the level of intensity and pointing noise. 
The fiber amplifier output is controlled by two feedback loops, which stabilize the light in low power and high power regimes, discussed later.
After intensity stabilization, the laser is mode-shaped to a desired waist.
The optical system is engineered to be passively stable in terms of pointing, but the position and phase of the optical lattice are monitored in real time, and beam pointing can be controlled with remotely actuated mirrors.

\section{Laser system}
\subsection{Seed laser system}
The lattice laser system is seeded by an Innolight Mephisto laser, which provides approximately $2\,$W of $1064\,$nm light. 
This commercial laser utilizes a non-planar ring oscillator (NPRO) design that ensures superb phase and intensity stability. 
This source is split into four fibers (Thorlabs PM980-XP), each connected to a fiber amplifier. 
Three of these fiber amplifiers are used to produce the 3D optical lattice~\cite{Parsons2016}, and a fourth is used for transport~\cite{Huber2014}.
To avoid interference of the lattice beams, two of the beam paths pass through acousto-optic modulators (AOM), which detuning the lattice frequencies from each other.

\subsection{Fiber amplifier system}
Quantum gas microscopy of \Li{} requires single site trap frequencies of approximately $1\,$MHz~\cite{Parsons2015} in order to be in the Lamb-Dicke regime necessary for Raman sideband cooling.
Such trap frequencies may only be attained using tightly focused, high-power beams. 
In practical terms, this requires approximately $30\,$W per lattice axis, which we attain using \textit{Nufern} Fiber amplifiers (NuAmp SUB1174-34).
These amplifiers can amplify $150\,$mW of seed light to up to approximately $45\,$W. They use a length of doped fiber pumped with $808\,$nm light from fiber coupled diode sources to amplify $1064\,$nm light. 
The amplification is performed using a two stage design, where the stages are controlled and powered using separate power supplies and electronics.

The high optical powers supplied by the fiber amplifiers raise two engineering difficulties:
\begin{enumerate}
  \item The fiber amplification process is susceptible to many technical noise sources, such as acoustic coupling of the fiber to the environment and electrical design of the amplifier, 
  \item Stimulates Brillouin scattering (SBS) places a hard lower bound on the noise floor~\cite{Agrawal2013}. 
\end{enumerate}

The NuAmp suffers from several technical problems. 
It is controlled via a USB interface, which complicates communications with an array of these devices.  
Worse yet, it is suspected of causing ground loops which add noise and interference to the analog systems. 
This added noise can be seen in figure \ref{fig:noises}, where the fiber-amplified light exhibits many more spurs in the intensity noise spectrum than the seed laser.
We solve this problem by replacing the USB control board with a custom-designed solution  that is compatible with our custom control system and engineered to avoid ground loops.

Further sources of technicl noise arise from the design of the amplifier. There are two power supplies in the fiber amplifier - a general purpose power supply (PSU) for the control electronics and the first, low-power amplification stage and a high power supply for the second, high-power stage. 
Despite the presence of a water-cooled cold-plate for the gain fiber and pump diodes, the high power PSU is air cooled by an internal fan.
This fan is expected to increase noise at acoustic frequencies and degrade pointing stability by increasing vibrations.
In order to reduce acoustic coupling and reduce switching power supply noise, both power supplies were removed from the fiber amplifier chassis, and the low power supply was replaced by a linear PSU (Acopian A24MT350M). 
Further, a high current line filter (MPE DS26387) was added to the high-power PSU (Lumina LDD-600-60-10-5VP/M).

These upgrades serve to suppress many of the noise spurs measured in the relative intensity noise (RIN) of the fiber amplifier, as seen in figure \ref{fig:noises}. 
The measurements in parts (a)-(d) were carried out on a single fiber amplifier, prior to and after the modifications. The initial and final noise spectra were verified to be consistent between amplifiers coming from the same manufacturing batch.
The figure is based on fiber amplifiers from a relatively new batch, which, even in its unmodified configuration does not exhibit a large switching supply spur. 
This is not the case for certain older batches, where the spur is approximately 30 dB higher than the background.

The SBS noise is fundamental to the fiber amplifier architecture and cannot be entirely avoided. However, since it is a function of fiber length and optical power, it can be minimized. To this end, we use custom-ordered amplifiers with a $0.5\,$m length output fiber, compared to the few meter standard lengths.
In addition, we set the lowest pump diode current that yields enough power to perform imaging. 
The rise of the noise floor as a function of the pump diode current can be seen in figure \ref{fig:noises}(a).

@@

The operation of the Nufern fiber amplifiers can be further optimized by finely controlling the cooling water.
The water temperature tunes the wavelength of the light supplied by the diodes that optically pump the gain fiber.
Optimizing the temperature is advantageous because (1) the maximum output power of the fiber amplifier is increased due to the improved efficiency of pump light conversion and (2) the lifetime of the amplifier is increased because less stray pump light needs to be absorbed when the pump light is separated from the output, reducing the thermal load.

This is possible because the pump diodes are fixed directly to the water-cooling plate without additional temperature regulation.
Adding a Peltier TEC and regulating to the optimal wavelength is possible, but technically challenging (requiring cooling power of $\approx100\,$W per each of the three pump diodes).
We use a single, dedicated heat exchanger (TermoTek P21518970) to supply cooling water to an array of four fiber amplifiers. 
We tune the water temperature such that the average power supplied by the amplifiers supplying the $X$ and $Y$ lattices is optimized, while simultaneously keeping all reported amplifier temperatures below $\leq 30^\circ\mathrm{C}$\footnote{As recommended by a \textit{Nuferm} support engineer.}.
Experimentally, we have found that more power can be gained by running the amplifiers at hotter temperature.

@@

\subsection{Optical system}
The free-space propagation of the fiber amplifier output presents several technical challenges and constraints:
\begin{enumerate}
  \item The high optical power poses a significant danger to users and equipment.
  \item The high optical power is sufficient to induce substantial thermal lensing in some optics. 
  \item The positional fluctuations of the waist at the atoms cannot be allowed to exceed a few microns between experimental realizations.
\item By construction, the beam is retro-reflected, threatening the lifetime of the amplifier. 
  \end{enumerate}

In order to address these challenges, we use a carefully engineered an optical system, shown in figure \ref{fig:optical_layout}, that collimates and polarizes the beam, controls its intensity between $10\,\mu$W and $20$ W, and ensures good pointing stability.
Due to the high laser power, fused silica glass and IBS coatings are used wherever possible.
Since reflections pose a danger to the users and equipment, all undesired beams above $100\,$ mW are directed to water cooled beam-dumps (to prevent temperature gradients on the table) placed as far as practical from critical beam paths.
Weaker beams are dumped on uncooled diffusive beam catchers.

The large-diameter fiber tip at the output of the fiber amplifier is mounted in a monolithic mount made from oxygen-depleted copper.
The beam is collimated using an $f=20\,$mm fused-silica, air-spaced triplet collimator (Opto-Sigma HFTLSQ-15-20PF1).
The polarization is set using an IBS coated thin-film plate polarizer (Precision photonics PO1056-FY), which has the added benefit of rejecting the undesired cladding modes of the fiber, which carry about $1\,$W of light. 
The fact that the lattice is retro-reflected, combined with the observation that optical isolation falls with applied optical power~\cite{Yoshida1999} means that two stages of optical isolators must be used. 
The optical isolators are based on $5$ mm diameter isolators from Thorlabs (IO-5-1064-VHP), with the default cube polarizers replaced by IBS coated thin-film plate polarizers. 
The isolators are located approximately $1\,$m away from the ultracold atoms which are susceptible to  undesired stray fields.
To overcome this problem the isolators are enclosed in mu-metal shielding.
Actuation of the power is performed after the optical isolators, meaning that the full fiber amplifier output is continually passed through them. 
This is advantageous because there are no thermalization transients introduced during the experimental sequence.

@@

After the beam has been cleaned up and isolated, it is important to consider two aspects of the desired experiment: 
\begin{enumerate}
\item The lattice power must be continuously tunable from around $10$ $\mu$W to around $20$ W. 
\item The location of the minimum Gaussian beam waist must be accurately positioned to overlap the other lattices and dipole traps present in the experiment, and must remain there for the duration of the collection process (hours to days).
\end{enumerate}
To satisfy the first of these requirements, a two-stage feedback system is implemented. 
Two stages are used because the experiment typically operates in one of two regimes, which we term \textit{detection} and \textit{interaction}. 
In the interaction phase of the experiment, the lattice is relatively shallow ($\approx 100\,$mW, corresponding to depths of $\approx 10 E_R$, where $E_R$ is the geometric recoil of the lattice), allowing atoms to tunnel and interact with each other. 
In this regime, fine and potentially fast control is required.
In the detection phase, the depth of the lattice is dramatically raised ($\approx 2000 E_R$), isolating atoms in their individual wells so that Raman sideband imaging can be performed~\cite{Parsons2015}.
In the detection regime, the control need not be fast, and the passive stability of the Mephisto laser ensures that noise is low at relevant frequency scales ($\approx 2$ MHz).
Thus, two control loops are utilized, a fast loop that uses an acousto-optic modulator to actuate the laser power at low powers, and a slow loop which uses a Berek compensator to actuate the laser in high power mode.

After the isolation stage, we use the sequence of optical elements shown in figure \ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on beam path) half-wave plate, and polarizer. 
The Berek compensator is a $z$-cut quartz plate coated to be anti-reflecting at $1064\,$nm. 
By tilting the plate about its vertical axis by a few degrees, the extraordinary axis is mixed into the propagation of the beam, leading to  tilt-dependent bi-refringence that can be tuned from a zero-wave plate past a half-wave plate, represented by a retardation angle $\phi$.
Combined with the subsequent polarization optics, rotation of the Berek compensator can vary the transmitted power with perfect contrast. 
Figure \ref{fig:berek} shows the transmitted power as a function of retrardation angle $\phi$ and fast-axis angle $\theta$ of the waveplates surrounding the compensator.
\begin{figure}
  \begin{center}
    \includegraphics{fig/berek_sim.pdf}
    \caption{\textbf{Berek compensator} performance as a function of the Berek tilt angle $\phi$ and waveplate tilt angles.}\label{fig:berek}
  \end{center}
\end{figure}

Full modulation contrast is achieved when $\theta=22.5^\circ$ to the rotation axis of the Berek compensator.
This contrast can be reduced by varying $\theta$, to the limiting case where $\theta=0$, where the rotation of the Berek compensator does nothing, since the polarization of the light is parallel to the rotation axis of the quartz plate.

\begin{figure*}
  \begin{center}
    \includegraphics{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop} performance showing the impulse response (a) at relatively low powers and the corresponding Bode plots (b-c).}\label{fig:berek_step_response}
  \end{center}
\end{figure*}
A key point, as seen in figure \ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator. 
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in figure \ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate. 
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.
Since two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously. 
To accomplish this, the low power loop is always manually railed to its high extreme when the high power loop is active, and the digital high power loop is disabled when the low power loop is active.

\subsection{Low power feedback}
The high power feedback used in the \textit{imaging} phase of the experiment is exceedingly robust because the heating rate from scattered resonant imaging light far exceeds any technical heating from the loop.
Even if the loop was exceedingly noisy, the heating would be entirely compensated by the simultaneous Raman cooling~\cite{Parsons2015}.

In contrast, the low power loop that controls the lattice laser during the \textit{interaction} phase of the experiment is critically important:
\begin{enumerate}
  \item Its RIN can limit the system temperature.
  \item It can affect the pointing.
  \item It must have closed loop bandwidths of $\geq1\mathrm{\,kHz}$, and open loop modulation bandwidths $\geq 100\mathrm{\,kHz}$.
\end{enumerate}
To this end, we use an all-custom loop based on a TeO$_2$ AOM (Gooch \& Housego AOMO 3080-198), where the power of the RF supplied to the AOM is used to actuate the fast, low power feedback loop.
The AOM was chosen for its low thermal lensing and easily-accessible RF powers. 
It is mounted on a custom-built, monolithic, water-cooled flexure mount that allows optimization of the AOM efficiency, that has been found to remain stable over years of operation.

\begin{figure}
  \begin{center}
    \includegraphics{fig/circuits.pdf}
    \caption{\textbf{Feedback circuits.} (a) The reverse-biased photodiode. (b) Externally referenced phase-locked loop. (c) Nonlinear loop filter. (d) RF power modulation using a mixer.}\label{fig:circuits}
  \end{center}
\end{figure}
The feedback circuit is shown in figure \ref{fig:circuits}a. As seen in figure \ref{fig:optical_layout}, a pair of low noise photodiodes measures the optical power of the beam sampled with a weakly reflective optic (Thorlabs BSF10-C). 
The beam sampler is wedged in such a way that the two reflections diverge from one another, allowing us to illuminate both photodiodes. 
One of the sampled beams is attenuated by an additional factor of $20\times$ by reflection off of an uncoated fused silica glass plate, such that the dynamic range of its photodiode covers $0-20$ W of the beam directed at the atoms, and controls the Berek compensator based, high-power, feedback loop. 
The other beam is not attenuated, save by the beam sampler, and its dynamic range covers $0-1$ W.
To prevent damage to the low power photodiode when the system is at its highest powers, a shutter (Stanford Research Systems SR475) blocks the light when the high power loop is in use.
The low power photodiode passes its signal $V_{\mathrm{pd}}$ to a loop filter circuit, which compares it to the control system's set-point, generated by digital to analog converter.
The loop filter's output controls the IF port of a microwave mixer (Minicircuits ZFM-2-S+), effectively modulating the RF power supplied by a low noise phase-locked loop (PLL).
The resulting RF signal is then amplified and sent to the AOM.
Let us now consider every part of this signal chain individually.

\subsubsection{Phase-locked loop (PLL)}
The local oscillator (LO) RF source is a custom-built printed circuit board (PCB) implementing a phase locked loop (PLL) based on the ADF4002 IC and Crystek's CVCO55CL-0060-0110 voltage controlled oscillator (VCO), preamplified with a low noise amplifier (Minicircuits PSA4-5043+). 
To ensure frequency stability (which translates directly into beam pointing of the laser), the PLL is locked to the $10$ MHz clock distributed around the lab.
In addition to PLL functionality, the board also contains two digitally-controlled, high-isolation RF switches, giving a total isolation exceeding $90$ dB.

The spectrum and relative intensity noise of the PLL can be found in figure \ref{fig:pll}. 
Vitally, the spectrum of the PLL exhibits phase noise comparable to (albeit worse than) that of a typical VCO, without the slow frequency drift characteristic to these devices. 
Elimination of the carrier drift is worth the added phase noise, however, because any drift of the frequency affects the pointing of the laser beam passing through the AOM, which would misalign the optical lattice.
